---
title: "Diabetes.markdown"
author: "nirman"
date: "2023-11-29"
output:
  html_document: 
    keep_md: yes
  pdf_document: default
---

Importing the packages and calling it by library function

```{r}
library(ggplot2)
library(reshape2)
library(imputeTS)  ##for mean imputation
library(caTools)    ##package for train test split
library(dplyr)
library(ROCR)
library(caret)
library(verification)
```

Loaded the dataset from local system

```{r}
df=read.csv("F:/Desktop items/datasets/diabetes.csv")
head(df)
tail(df)
```

Checking data types for all variables

```{r}
str(df)

```

```{r}
sum(is.na(df))
```

No missing values

```{r}
sum(duplicated(df))
```

No duplicate values

Checking for inappropriate or zero values in variables

```{r}
for (col in names(df)) {
  zero_values <- sum(df[[col]] <= 0)
  cat(paste('Zero values in column', col, '=', zero_values, '\n'))
}

```

The BMI, Insulin, Skin thickness, Blood pressure and Glucose cannot have zero values so imputing it with the median values

```{r}
col_containing_zero <- c("Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI")

for (col in col_containing_zero) {
  median_value <- median(df[[col]], na.rm = TRUE)
  df[df[[col]] <= 0, col] <- median_value
}
```

Again checking zero values after imputation

```{r}
for (col in names(df)) {
  zero_values <- sum(df[[col]] <= 0)
  cat(paste('Zero values in column', col, '=', zero_values, '\n'))
}
```

EXPLORATORY DATA ANALYSIS \\

Comparison of patients with and without diabetes through Bar charts

```{r}
df %>%
  group_by(Outcome)%>%
  summarise(n=n())%>%
  ggplot(aes(x=Outcome,y=n))+
    geom_bar(stat = 'identity',fill='purple')+
    geom_label(aes(label=n))+
  theme_minimal()+
  labs(title = "Count of Diabetic patients and Non-Diabetic patients")+
  theme(panel.grid = element_blank())+
  theme(plot.title = element_text(hjust = 0.5))
```

Comparing different variables with diabetic and non diabetic patients

```{r}
dist_columns <- c('Pregnancies', 'Glucose', 'BloodPressure', 'SkinThickness', 'Insulin', 'BMI', 'DiabetesPedigreeFunction', 'Age')

for (col in dist_columns) {
  ggplot(df, aes(x = !!sym(col))) +
    geom_histogram(binwidth = 1, fill = 'blue', color = 'black', alpha = 0.7) +
    geom_density(color = 'red') +
    labs(title = paste("Distribution of", col, "column")) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    scale_fill_manual(values = 'blue') +
    scale_color_manual(values = 'black') +
    facet_wrap(~1, scales = 'free_y') +
    NULL
}
```

Obtaining correlations matrix

```{r}

cor(df)   ##computed correlation matrix 
corr_mat=round(cor(df),2)  ##rounded to 2 decimel
melted_corr_mat <- melt(corr_mat)           #reduce the size of correlation matrix
```

Plotting the correlation heatmap with annotations

```{r}
ggplot(data = melted_corr_mat, aes(x=Var1, y=Var2,
                                   fill=value)) + 
geom_tile()+
geom_text(aes(Var2, Var1, label = value), 
          color = "black", size = 4)
```

looking for missing values

```{r}
 
sum(is.na(df$Pregnancies))
sum(is.na(df$Glucose))
sum(is.na(df$BloodPressure))
sum(is.na(df$SkinThickness))
sum(is.na(df$Insulin))
sum(is.na(df$BMI))
sum(is.na(df$Age))
sum(is.na(df$Outcome))
#so the data has no missing values
```

Checking Outliers using Scatterlot

```{r}
plot(df$Glucose)
plot(df$BloodPressure)
plot(df$SkinThickness)
plot(df$Insulin)
plot(df$BMI)
plot(df$Age)
```

it is observed that the above features have many outliers and the outliers are zero values \####

```{r}
###check for summary of data frame
summary(df)
colSums(df==0)  ##checking the number of zeros in each features
```

```{r}
#replacing the zeros by imputing mean in each of the important features
##converting the zero values to NA 
df$Glucose[df$Glucose==0]<-NA
df$BloodPressure[df$BloodPressure==0]<-NA
df$SkinThickness[df$SkinThickness==0]<-NA
df$Insulin[df$Insulin==0]<-NA
df$BMI[df$BMI==0]<-NA

df$Glucose<-round(na_mean(df$Glucose, option = "mean", maxgap = Inf),0)##this is only for mean imputation
df$BloodPressure<-round(na_mean(df$BloodPressure, option = "mean", maxgap = Inf),0)##this is only for mean imputation
df$SkinThickness<-round(na_mean(df$SkinThickness, option = "mean", maxgap = Inf),0)##this is only for mean imputation
df$Insulin<-round(na_mean(df$Insulin, option = "mean", maxgap = Inf),0)##this is only for mean imputation
df$BMI<-round(na_mean(df$BMI, option = "mean", maxgap = Inf),1)##this is only for mean imputation

```

Logistic Regression Model

```{r}
##split features and target 
X=df[-8]          ##features
y=df[8]		        ##target

```

Splitting Train Test Data

```{r}
##splitting the data into trainning and test data

set.seed(1)
sample <- sample.split(df, SplitRatio = 0.8)
train_df  <- subset(df, sample == TRUE)
test_df   <- subset(df, sample == FALSE)
```

```{r}
##spliting the data set into outcome and features
train_X=train_df[-8]
train_y=train_df[8]
test_X=test_df[-8]
test_y=test_df[8]
```

```{r}
#checking the dimensions  
dim(train_y)
dim(test_y)
dim(train_X)
dim(test_X)
head(train_X)
head(train_df)
```

```{r}
###Model training
logit_model<- glm( Outcome ~  Age+BMI+ Insulin+SkinThickness+BloodPressure+Glucose+Pregnancies,
                    data = train_df,
                    family = "binomial")
summary(logit_model)
```

```{r}
###model testing
predict_model <- predict(logit_model,
                       test_df, type = "response")
```

```{r}
###model performance
test_class<-ifelse(predict_model<=0.5, 0, 1)   ###assuming 0.5 as the optimal threshold probability
test_class
```

```{r}
#it is required to  factor the outcomes in order to create a confusion matrix
test_class<-as.factor(test_class)
test_df$Outcome<-as.factor(test_df$Outcome)

confusionMatrix(test_df$Outcome,test_class)
```
